Risk-neutral density extraction procedure
========================================================

```{r setoptions, echo = FALSE}
# setwd('Desktop/workspace/extractRnd/demo/')
opts_chunk$set(echo = TRUE, results='markup', cache = TRUE)
```

```{r session}
sessionInfo()
```

I demo how to extract risk-neutral density from the implied volatility surface. Data comes from optionmetrics. Extraction method heavily follows Figlewski 2010. 

## Read in data. 

Sample data is the daily vol surface of IBM and MSFT during August of 2013. 

```{r}
data = read.csv('sampledata_optionmetrics.csv')
data = data[,c('date','days','ticker','cp_flag','impl_strike','delta','impl_volatility')]   # these variables are relevant
summary(data)
```

Let's visualize the 1-quarter-ahead vol curves for IBM on the first day of Aug 2013. Note that puts are much more expensive than calls (does this not violate put-call parity?)

```{r}
# {r, fig.width=7, fig.height=6}
subdata = subset(data, (ticker == 'IBM') & (date == min(data$date)) & (days == 91))
with(subset(subdata, cp_flag == 'C'), plot(impl_strike, impl_volatility, col='red', xlab = 'implied strike price', ylab='implied volatility', main = 'IBM, 1 quarter ahead'))
with(subset(subdata, cp_flag == 'P'), points(impl_strike, impl_volatility, col='blue'))
legend("topright", pch = 1, col = c("red", "blue"), legend = c("call", "put"))
```

